---
title: "Using `ddcurves2` to fit density profile models"
author: "Andrew Manderson"
date: "`r format(Sys.time(), '%d %B, %Y')`"
fontfamily: kpfonts
fontsize: 10pt
papersize: a4
geometry: margin=1.25cm
output: 
  pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, comment = NA, eval = TRUE)
```

`ddcurves2` is an `R` package for fitting the hierarchical model detailed
in the forthcoming publication "Statistical estimation of temporal variability
in vertical seawater density profiles". It also has another model for 
temporal variation of background density profiles, that may be discussed at a 
later date.

## Overview of this document

There are three topics this document covers:

* The dependencies required by `ddcurves2`, some basic instructions on 
installing them, and where to documentation in case of installation difficulties.
* Fitting the hierarchical model to the data presented in the publication
* Demonstrating some of the built in plotting functionality inside the package.

## Installation

#### `R`

You will need an installation of `R` (https://www.r-project.org/), and I 
recommend using RStudio (https://www.rstudio.com/) to run `R`.
Both of these websites have installation documentation should you run into
difficulties.

#### `R` packages

There are a number of `R` packages we require, which we can install with
`install.packages()`. For now, we require `devtools`, `ggplot2` and `Rcpp`, 
which can be installed within `R` with the following command:

```{r dep_package, eval = FALSE}
install.packages(c("devtools", "ggplot2", "Rcpp"))
```


#### `Stan` and `rstan`

This package makes use of `Stan` to fit the models. Installing it can be somewhat
of a pain. The documentation for installing `Stan` and its dependencies is fairly
comprehensive, and is available for either Windows or OSX / Linux at:

* Windows: https://github.com/stan-dev/rstan/wiki/Installing-RStan-on-Windows
* OSX / Linux: https://github.com/stan-dev/rstan/wiki/Installing-RStan-on-Mac-or-Linux

If the OSX installation instructions do no work for you, I can recommend running
`xcode-select --install` from the terminal to get the `c++` compiler (often the
most error prone step), or checking the Stan forums  
(http://discourse.mc-stan.org/) for similar issues.


#### `ddcurves2`

Finally we can install `ddcurves2`, which makes use of 
`devtools::install_github()` to install the most recent version off of github.

``` {r inst_package, eval = FALSE}
devtools::install_github("hhau/ddcurves2") # should be public by the time this goes out.
```

#### `iwaves`

To make some of the figures in the paper, we make use of the python package
iwaves: (https://bitbucket.org/mrayson/iwaves). Installation instructions 
can be found on the bitbucket page. 

## Running the hierarchical model

The data presented in the paper is included in the package. It is split into
the two halves. We fit it as two separate models, as it makes minimal difference
to the estimated posterior distributions of each curve, and the data has 
a variable number of depths at which density data is recorded. 

For the first half of the data, we load the library and create the list of 
data to give to the fitting function, and call the fitting function with
some additional options to `Stan`. The model requires feasible initial values in
order to function appropriately, which we supply.

```{r first_half_data}
library(ddcurves2)

stan_dat_one <- list(
  n_depths = length(Crux_KP150_Phs1$depths),
  n_times = nrow(Crux_KP150_Phs1$density_mat),
  depths = Crux_KP150_Phs1$depths,
  densities = Crux_KP150_Phs1$density_mat
)

# model needs initial values
n_t_one <- stan_dat_one$n_times
init_sub_one <- list(
  beta_zero = rep(1025, n_t_one),
  beta_one = rep(1, n_t_one),
  beta_three = rep(42, n_t_one),
  beta_six = rep(48, n_t_one),
  beta_midpoint = array(c(rep(77, n_t_one), rep(150, n_t_one)), dim = c(n_t_one, 2)),
  mean_beta_zero = 1025,
  mean_beta_one = 1,
  mean_beta_three = 42,
  mean_beta_six = 48,
  mean_beta_midpoint = array(c(77, 150), dim = 2),
  sigma_beta = array(c(0.18, 0.12, 26.1, 8.09, 12.6, 6.65), dim = 6),
  sigma_curve = 0.08
)

model_fit_one <- double_tanh_no_timeseries(
  stan_data_list = stan_dat_one,
  iter = 1500,
  warmup = 1000,
  save_warmup = FALSE,
  chains = 3,
  cores = 3,
  refresh = 25,
  control = list(adapt_delta = 0.95, max_treedepth = 16, stepsize = 1e-6),
  init = list(init_sub_one, init_jitter(init_sub_one), init_jitter(init_sub_one))
)
saveRDS(model_fit_one, file = "model-fit-one.rds")
```


And the same for the second half of the data: 

```{r sec_half_data}
stan_dat_two <- list(
  n_depths = length(Crux_KP150_Phs2$depths),
  n_times = nrow(Crux_KP150_Phs2$density_mat),
  depths = Crux_KP150_Phs2$depths,
  densities = Crux_KP150_Phs2$density_mat
)

n_t_two <- stan_dat_two$n_times
init_sub_two <- list(
  beta_zero = rep(1025, n_t_two),
  beta_one = rep(1, n_t_two),
  beta_three = rep(42, n_t_two),
  beta_six = rep(48, n_t_two),
  beta_midpoint = array(c(rep(77, n_t_two), rep(150, n_t_two)), dim = c(n_t_two, 2)),
  mean_beta_zero = 1025,
  mean_beta_one = 1,
  mean_beta_three = 42,
  mean_beta_six = 48,
  mean_beta_midpoint = array(c(77, 150), dim = 2),
  sigma_beta = array(c(0.18, 0.12, 26.1, 8.09, 12.6, 6.65), dim = 6),
  sigma_curve = 0.08
)

model_fit_two <- double_tanh_no_timeseries(
  stan_data_list = stan_dat_two,
  iter = 2500,
  warmup = 2000,
  save_warmup = FALSE,
  chains = 3,
  cores = 3,
  refresh = 25,
  control = list(adapt_delta = 0.95, max_treedepth = 16, stepsize = 1e-6),
  init = list(init_sub_two, init_jitter(init_sub_two), init_jitter(init_sub_two))
)
saveRDS(model_fit_two, file = "model-fit-two.rds")
```

```{r, echo=FALSE}
## for use later - in other figures
# saveRDS(model_fit_one, file = "model-fit-one.rds")
# saveRDS(model_fit_two, file = "model-fit-two.rds")
# model_fit_one <- readRDS("./model-fit-one.rds")
# model_fit_two <- readRDS("./model-fit-two.rds")
```


## Some output plots

#### Estimated density profiles

For a given time index, we can look at the estimated density profile:

```{r, fig.height = 8, fig.width = 14}
t_index <- 123

dens_plot_one <- ddcurves2::double_fitted_curve_plotter(
  model_fit_one,
  t_index,
  stan_dat_one
)

dens_plot_two <- ddcurves2::double_fitted_curve_plotter(
  model_fit_two,
  t_index,
  stan_dat_two
)

library(gridExtra)

# put both plots together
grid.arrange(arrangeGrob(
  dens_plot_one,
  dens_plot_two,
  nrow = 1
))

```

Here we are looking at the 123rd observation from the start of the first and
second halves of the overall data set. 

We can also look at the distribution of the means of the parameters of the curve,
for which we will only plot the first model:

```{r, fig.height = 8, fig.width = 14}
mean_beta_density_diag(model_fit_one)
mean_beta_density_diag(model_fit_two)
```

For more examples of the R package, as well as how we use the python package,
we direct the reader to the code that produces the figures in the paper.

 
 
